---
title: 'Assignment 3: Data Exploration'
author: 'Suad Muradov, Section #3'
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
getwd()
Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors=TRUE)
Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

Litter$collectDate<-as.Date(Litter$collectDate,"%Y-%m-%d")
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Neonicotinoids are used for protecting the farmer plants and crops by killing the insects. These substances can lead to the decrease in the bee populations which would reduce the pollination of plants, cause natural imbalance and global food shortages.

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Woody waste includes logs, branches, dead trees, and root wads that are all dead woody plant detritus. As a source of energy and shelter for terrestrial and aquatic animals, woody debris is an essential component of forest ecosystems. Woody debris also influences sediment transport by changing water flow shape and roughness, which in turn affects water flow and sediment movement. Such properties of woody debris helps us understand the nature of environmental balance and biodiversity that later helps us to construct policies for the protection of forests.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*
*
*



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics) #4623 rows and 30 columns
```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect) 
```

> Answer: Most common effects are those of effects on *population*, *mortality* and *behavior*. These effects of neonicotinoids are crucial for us because we want to know what is the specific effects of using these insecticides on the dynamics of group populations, to see how fast their populations decrease and on top of that we are interested in looking at how resistant these pests are to the neonicotinoids (do their effect make them evolve or are they overly effective killers).

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics)
```

> Answer: All of these species are the parts of the same groups of bees. As we mentioned above, bees are more important species to humans than any other pests from whom we are trying to protect our plants. Other insects are not particularly harmful to the global food chain and therefore, such research is employed to understand the extent of harm these neonicotinoids have on bees so that later we could evolve these substances into ones that kill all of the rest of the pests, without harming bees.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
```

> Answer: It is common to refer to categorical variables in R as factors because they can take both ordered and unordered discrete values. Conc.1..Author. is a factor with 1006 pre-defined set values, looks like character vector but it is actually an integer. Because categorical variables are treated differently from continuous variables in statistical models, factors can be easily used for statistical analysis and for plotting.

## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r, fig.height=4, fig.width=5}
library(ggplot2)
ggplot(Neonics) + geom_freqpoly(aes(x = Publication.Year), bins = 10)
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color=Test.Location), bins = 10)
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r, fig.height=4, fig.width=7}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar()
```

> Answer: Lowest-observable-effect-level and No-observable-effect-level for Terrestrial species were the two most common endpoints. 
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
library(lubridate)
class(Litter$collectDate)
factor2date<-ymd(Litter$collectDate)
class(factor2date)
unique(factor2date,incomparables = FALSE)

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID,incomparables = FALSE)
summary(Litter$plotID)

```

> Answer: While 'summary' function is designed to tell us which properties is included into our model, it does not show us the observations. 'unique' function removes duplicate elements and rows from an array, data frame, or vector similar to x and then shows the observations

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: 